Relationships in the extremes and their influence on competition and coexistence

### 2022-07-10 paper/Paper.Rnw
some more edits

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it extinct \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``relative nonlinearity
in competition'' and ``storage effect'' \citep{Chesson_1994} which will be familiar to many readers. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2006, Adler_2010}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Caceres_1997, Descamps_2005, Narawani_2013}, and corals \citep{Alvarez-Noriega_2020}. 
Although MCT is mathematically complex to many ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

@@

%Lottery model and results based on it
To further demonstrate the plausibility of ATAs altering coexistence between species, we carried out simulations using a simple two-species model, the lottery model. 
The simulations described in this paragraph are sufficient to demonstrate that ATAs influence coexistence in some manner; 
the precise nature of that influence is then the subject of the rest of the paper.
The lottery model was originally introduced as a model for coral reef fish competing for
a set of $N$ breeding territories.
Letting $N_i(t)$ denote the population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is a mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that adults die at rate $\delta$ at each time step, and are replaced by new juveniles 
in proportion to the reproductive outputs of the two species that year. Note that $N$ is constant through time. 
For each $i$, we assume for simplicity that the random variables $B_i(t)$ are independent and identically distributed (iid) through
time. We let $B_i = \exp(b_i)$, where $\left(b_1, b_2\right)$ is some bivariate random variable with the property 
that $b_i$ is normally distributed with mean $\mu_i$ and variance $\sigma^2$. 
We do \emph{not} assume that $\left(b_1, b_2\right)$ is a 
bivariate normal distribution. 
For proof of concept, we simulated two communities for \Sexpr{M} time 
steps each, using symmetric and asymmetric tail associations (Fig. \ref{fig:noise}) between $b_1$ and $b_2$, respectively, in the
two simulations. Fig. \ref{fig:pedagog} gives details of the 
simulations and shows that ATAs substantially influenced aspects of species coexistence for this model setup.

>>>>

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it extinct \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``relative nonlinearity
in competition'' and ``storage effect'' \citep{Chesson_1994} which will be familiar to many readers. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2006, Adler_2010}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Caceres_1997, Descamps_2005, Narawani_2013}, and corals \citep{Alvarez-Noriega_2020}. 
%Although MCT is mathematically complex to many ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.
Though MCT's original analytical form presented mathematical challenges for applications to and tests of new systems, recent extensions expand testability and accessibility \citep{Ellner_2016,Ellner_2019}.

@@

%Lottery model and results based on it
%To further demonstrate the plausibility of ATAs altering coexistence between species, 
To more robustly demonstrate the role of ATAs in structuring fluctuation-dependent coexistence, we carried out simulations using a simple two-species model, the lottery model. 
The simulations described in this paragraph are sufficient to demonstrate that ATAs influence coexistence in some manner; 
the precise nature of that influence is then the subject of the rest of the paper.
The lottery model was originally introduced as a model for coral reef fish competing for
a set of $N$ breeding territories.
Letting $N_i(t)$ denote the population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is a mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that adults die at rate $\delta$ at each time step, and are replaced by new juveniles 
in proportion to the reproductive outputs of the two species that year. Note that $N$ is constant through time. 
For each $i$, we assume for simplicity that the random variables $B_i(t)$ are independent and identically distributed (iid) through
time. We let $B_i = \exp(b_i)$, where $\left(b_1, b_2\right)$ is some bivariate random variable with the property 
that $b_i$ is normally distributed with mean $\mu_i$ and variance $\sigma^2$. 
We do \emph{not} assume that $\left(b_1, b_2\right)$ is a 
bivariate normal distribution. 
For proof of concept, we simulated two communities for \Sexpr{M} time 
steps each, using symmetric and asymmetric tail associations (Fig. \ref{fig:noise}) between $b_1$ and $b_2$, respectively, in the
two simulations. Fig. \ref{fig:pedagog} gives details of the 
simulations and shows that ATAs substantially influenced aspects of species coexistence for this model setup.
